Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka

### 2025-08-14 analysis/paper.qmd
Update results and discussion

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  The models performance in predicting the occurrence of specific species at 
  archaeological sites is highly variable, but in aggregate the predictions are
  coherent and align with broad-scale trends in the ubiquity and relative 
  abundance of species in the archaeological record.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
---

@@

```{r dependencies, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")      )

@@

    filter(n_archaeo >= min_n_archaeo)

@@

```{r data-background}
# TODO: slow - cache
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_eurasia_land, n = n_background),
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
```

@@

if (!file_exists(derived_data("flora_background.Rdata"))) {
  flora <- flora |> 
    mutate(
      occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, remove = FALSE),
      occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
      occ = map(occ, bind_extract, st_as_stars(terrain)),
      occ = map(occ, bind_extract, st_as_stars(soil))
    )
  saveRDS(flora, derived_data("flora_background.Rdata"))
}
flora <- readRDS(derived_data("flora_background.Rdata"))
```

@@

```{r niche-data}
# Split training and test sets for model assessment 
flora <-  flora |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
  )

@@

```{r niche-fit}
# Fit best model to training data
if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
  flora <- flora |>
    mutate(
      niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
    )
  saveRDS(flora, derived_data("flora_niche_fit.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_fit.Rdata"))
```

<!-- TODO: prediction, hindcasting and hindcasting assessment methodology -->

```{r niche-predict}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  # flora <-
  flora |>
    mutate(
      pred = map(niche, collect_predictions)
    )

      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = env_w_asia, .progress = TRUE),
      # Binarize
      # TODO: use max SSS
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
      palaeodist = map(palaeodist, mutate, absent = prediction.FALSE. > 0.5)
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))

@@

```{r data-model-assessment}
flora <- mutate(
  flora,
  niche_test_metrics = map(niche, collect_metrics),
  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
                    event_level = "second"), # reference present=TRUE
  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
) |>
  select(-sensitivity)

@@

```{r fig-model-sensitivity-vs-n_occ, warning=FALSE}
#| fig-cap: Model sensitivity by number of training occurrences
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "sens") |>
  mutate(label = if_else(.estimate < 0.85, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x)) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic", size = 10/.pt) +
  scale_x_log10(name = "Occurrences") +
  scale_y_continuous(name = "Sensitivity", labels = label_percent()) +
  theme_minimal_grid() 
```

@@

```{r niche-predict-archaeo}
# Extract predictors and make predictions for archaeobotanical occurrences
# TODO: what about negatives?
if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
  archaeo_pred <- flora |>
    unnest(archaeo) |>
    nest(.by = c(climate_period), .key = "archaeo") |>
    mutate(
      archaeo = map(archaeo, st_as_sf, coords = c("longitude", "latitude"),
                    crs = latlong, remove = FALSE),
      archaeo = map2(archaeo, climate_period, function(x, period) {
        bind_extract(x, slice(climate, "period", as.character(period)))
      }),
      archaeo = map(archaeo, bind_extract, terrain),
      archaeo = map(archaeo, bind_extract, soil)
    ) |>
    unnest(archaeo) |>
    nest(.by = any_of(colnames(flora)), .key = "archaeo") |> # Restore original nesting
    mutate(
      archaeo_pred = map2(niche, archaeo, function(x, y) {
        # TODO: shouldn't be necessary - remove from workflow
        y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
        augment(extract_workflow(x), new_data = y)
      })
    ) |>
    select(taxon, archaeo_pred)

@@

```{r niche-pred-range}
area_present <- function(pred, res = pred_res) {
  st_apply(pred, "period", sum, na.rm = TRUE)["present"] |>
    data.frame() |>
    mutate(area_present = present * (pred_res)^2) |>
    select(period, area_present)
}

@@

# TODO:
# - alt. taxons as footnotes
# - check boxes of which (climate?) periods observed in ?
results_summary <- flora_all |>
  left_join(flora, join_by(taxon, taxon_aliases, archaeo, n_archaeo)) |>
  left_join(archaeo_pred_summary, join_by(taxon)) |>
  mutate(n_occ = coalesce(n_occ.y, n_occ.x), .before = n_occ.x) |>
  # Extract model metrics
  unnest(c(niche_test_metrics), keep_empty = TRUE) |>
  select(-.estimator, -.config) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  # Excluded samples
  mutate(exclude = (n_occ < min_n_occ)) |>
  # Prepare table
  select(taxon, taxon_aliases, n_archaeo, n_occ, exclude, accuracy, roc_auc, 
         sens, p_predicted) |>
  arrange(taxon)

results_summary |>
  select(-taxon_aliases) |>
  gt() |>
  cols_label_with(fn = str_to_sentence) |>
  cols_label(
    n_archaeo = "Arch.",
    n_occ = "Cur.",
    accuracy = "Acc.",
    roc_auc = "ROC-AUC",
    sens = "Sens.",
    p_predicted = "Sens. (arch.)"
  ) |>
  tab_spanner("Occurrences", columns = c(n_archaeo, n_occ)) |>
  tab_spanner("Model", columns = accuracy:p_predicted) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  add_taxon_alias_footnotes(results_summary$taxon_aliases) |>
  cols_align("right") |>
  cols_align("left", c(taxon)) |>
  fmt_number(accuracy:p_predicted) |>
  sub_missing() |>
  tab_style(cell_text(style = "italic"), cells_body(taxon)) |>
  tab_options(
    # footnotes.marks = "*",
    table.font.size = 11,
    table.width = pct(100),
    latex.use_longtable = TRUE
  )
```

@@

```{r fig-palaeodist-sum}
#| fig-cap: Predicted species richness (sum of predicted ranges) by period
# TODO: expand period facet labels
ggplot() +
  facet_wrap(vars(factor(period, levels = climate_periods$code, labels = climate_periods$label))) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_sum, present), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "Richness", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

@@

Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
Were these area part of the same broader 'interaction sphere' that produce Neolithic agriculture in West Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Western Anatolian (@fig-palaeodist-chickpea-rye).
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2001; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

```{r fig.height=10, fig-palaeodist-barely-levant, cache=TRUE}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
# TODO: why so slow?
palaeodist_hordeum <- do.call(c, c(
  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
))

ggplot() +
  facet_grid(taxon ~ fct_relabel(period, toupper)) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  scale_x_continuous(breaks = c(34, 36, 38)) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(
    xlim = c(33, 39), ylim = c(30, 38),
    crs = w_asia_albers, default_crs = latlong
  ) +
  labs(fill = "P(present)", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

@@

```{r fig-palaeodist-bread}
#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
bread <- c("Aegilops tauschii", "Triticum aestivum", 
           "Triticum turgidum dicoccum")
palaeodist_bread <- do.call(c, c(
  filter(flora, taxon %in% bread)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
)) |>
  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
  mutate(present = factor(present))

@@

## Hindcast models do not predict archaeobotanical composition {#sec-discuss-hindcasting}

The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
Since they do accurately predict the test dataset<!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.

For a variety of reasons, our models almost certainly underestimate the *fundamental* niche or potential distribution of the target taxon.
These include:

* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
Discrepencies between the modelled distribution, whether on the broad scale (e.g. the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e. false positives and false negatives), suggest several avenues for future investigation.

>>>>

---
title: Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  url: https://joeroe.io
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first continuous, spatially-explicit reconstructions 
  of the palaeodistributions of 82 plant species found regularly in association 
  with early agricultural archaeological sites in West Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model of each species based on its present-day distribution 
  in relation to climate and environmental variables. Predictions of the 
  potential ranges of these species at key stages of the Pleistocene–Holocene 
  transition could then be derived from these models using downsampled data from 
  palaeoclimate simulations.
  <!-- TODO: rewrite following reviews. Highlight BA to EH decline vs. 'amelioration' concept -->
  The models predict significant reductions and/or shifts in species ranges in
  the terminal Pleistocene and Early Holocene compared to present conditions. 
  Many species that are found throughout the region's 'hilly flanks' today 
  are indicated to have much more restricted distributions centered on the 
  Levant, Cyprus and Western Anatolia. In addition, overall ranges shrunk by an 
  average of c. 25% from the terminal Pleistocene to the Early Holocene. 
  The models performance in predicting the occurrence of specific species at 
  archaeological sites is highly variable, but in aggregate the predictions are
  coherent and align with broad-scale trends in the ubiquity and relative 
  abundance of species in the archaeological record.<!--
  however [...]
  The regional ubiquity of species in the archaeological record is [not] correlated with the 
  predicted size of its range and the diversity of archaeobotanical assemblages is [not] 
  correlated with the predicted diversity of its environs. This indicates that 
  trends in taxonomic composition of the archaeobotanical record is [not] likely 
  to have been influenced by environmental change and species turnover, in
  addition to human economic choices.-->
format: 
  elsevier-pdf:
    keep-tex: true
    # cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
execute: 
  echo: false
  cache: true
---

@@

```{r dependencies, cache=FALSE, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("probably")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"))

@@

    filter(
      n_archaeo >= min_n_archaeo | 
        taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")

@@

```{r data-background}
# TODO: slow - cache
if (!file_exists(derived_data("flora_background.Rdata"))) {
flora <- flora |>
  mutate(
    occ = map(occ, mutate, present = TRUE, weight = 1),
    bg = map(occ, function(x, region, n) { 
      background_sample(
        region = region, 
        N = n,
        coord_x = "longitude", coord_y = "latitude",
        present = FALSE,
        weight = (1 / n) * nrow(x) # weight presences and background equally
      )
    }, region = w_eurasia_land, n = n_background),
    occ = map2(occ, bg, bind_rows)
  ) |>
  select(-bg)
  saveRDS(flora, derived_data("flora_background.Rdata"))
}
flora <- readRDS(derived_data("flora_background.Rdata"))
```

@@

# TODO: slow, cache
flora <- flora |> 
  mutate(
    occ = map(occ, st_as_sf, coords = c("longitude", "latitude"), crs = latlong, 
              remove = FALSE),
    occ = map(occ, bind_extract, st_as_stars(hack(climate, "period", "cur"))),
    occ = map(occ, bind_extract, st_as_stars(terrain)),
    occ = map(occ, bind_extract, st_as_stars(soil))
  )
```

@@

```{r niche-data}
# Split training and test sets for model assessment 
flora <-  flora |>
  # Manually impute some problematic cases
  # mutate(
  #   occ = case_when(
  #     taxon_key %in% c(12225945, 7405743) ~ map(occ, function(x) {
  #       mutate(x, across())
  #     })
  #   )
  # ) |>
  mutate(
    # De-sf
    occ = map(occ, \(x) select(tibble(x), -geometry)),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE)))),
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
  )

@@

```{r niche-fit, cache.lazy=FALSE}
# Don't use lazy cache here: https://github.com/yihui/knitr/issues/572
# Fit best model to training data
if (!file_exists(derived_data("flora_niche_fit.Rdata"))) {
  flora <- flora |>
    mutate(
      niche = map(occ_split, \(x) last_fit(niche_workflow, x), .progress = TRUE)
    )
  saveRDS(flora, derived_data("flora_niche_fit.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_fit.Rdata"))
```

```{r niche-predict, cache.lazy=FALSE}
# NOTE: high memory usage
if (!file_exists(derived_data("flora_niche_predictions.Rdata"))) {
  flora <- flora |>
    mutate(
      pred = map(niche, collect_predictions),
      palaeodist = map(niche, \(mod, obj) predict(obj, extract_fit_engine(mod)),
                       obj = env_w_asia, .progress = TRUE),
      # Binarize
      palaeodist = map(palaeodist, mutate, present = prediction.TRUE. > 0.5),
    )
  saveRDS(flora, derived_data("flora_niche_predictions.Rdata"))
}
flora <- readRDS(derived_data("flora_niche_predictions.Rdata"))

@@

```{r data-model-assessment, cache.lazy=FALSE}
flora <- mutate(
  flora,
  niche_test_metrics = map(niche, collect_metrics),
  sensitivity = map(pred, sens, truth = present, estimate = .pred_class,
                    event_level = "second"), # reference present=TRUE
  niche_test_metrics = map2(niche_test_metrics, sensitivity, bind_rows)
) |>
  select(-sensitivity)

@@

```{r fig-model-sensitivity-vs-n_occ, warning=FALSE}
#| fig-cap: Model sensitivity by number of training occurrences
flora |>
  unnest(niche_test_metrics) |>
  filter(.metric == "sens") |>
  mutate(label = if_else(.estimate < 0.50, taxon, NA_character_)) |>
  ggplot(aes(n_occ, .estimate)) + 
  geom_smooth(method = "glm", formula = y ~ log(x), fullrange = TRUE, 
              se = FALSE) +
  geom_point() + 
  geom_text_repel(aes(label = label), fontface = "italic", size = 10/.pt) +
  scale_x_log10(name = "Occurrences", labels = label_number()) +
  scale_y_continuous(name = "Sensitivity", labels = label_percent(),
                     limits = c(0, 1)) +
  theme_minimal_grid() 
```

@@

```{r niche-predict-archaeo, cache.lazy=FALSE}
# Extract predictors and make predictions for archaeobotanical occurrences
# TODO: what about negatives?
if (!file.exists(derived_data("archaeo_pred.Rdata"))) {
  archaeo_pred <- flora |>
    unnest(archaeo) |>
    # Add dummy GBIF id
    mutate(gbif_id = 0) |>
    nest(.by = c(climate_period), .key = "archaeo") |>
    mutate(
      archaeo = map(archaeo, st_as_sf, coords = c("longitude", "latitude"),
                    crs = latlong, remove = FALSE),
      archaeo = map2(archaeo, climate_period, function(x, period) {
        bind_extract(x, st_as_stars(hack(climate, "period", as.character(period))))
      }),
      archaeo = map(archaeo, bind_extract, st_as_stars(terrain)),
      archaeo = map(archaeo, bind_extract, st_as_stars(soil))
    ) |>
    unnest(archaeo) |>
    nest(.by = any_of(colnames(flora)), .key = "archaeo") |> # Restore original nesting
    mutate(
      archaeo_pred = map2(niche, archaeo, function(x, y) {
        # TODO: shouldn't be necessary - remove from workflow
        y <- mutate(y, gbif_key = NA, scientific_name = NA, species = NA)
        augment(extract_workflow(x), new_data = y)
      })
    ) |>
    select(taxon, archaeo_pred)

@@

```{r niche-pred-range, cache.lazy=FALSE}
area_present <- function(pred, res = pred_res) {
  st_apply(pred, "period", sum, na.rm = TRUE)["present"] |>
    data.frame() |>
    mutate(area_present = present * (pred_res)^2) |>
    select(period, area_present)
}

@@

# TODO:
# - alt. taxons as footnotes
# - check boxes of which (climate?) periods observed in ?
results_summary <- flora_all |>
  left_join(flora, join_by(taxon, taxon_aliases, archaeo, n_archaeo)) |>
  left_join(archaeo_pred_summary, join_by(taxon)) |>
  mutate(n_occ = coalesce(n_occ.y, n_occ.x), .before = n_occ.x) |>
  # Extract model metrics
  unnest(c(niche_test_metrics), keep_empty = TRUE) |>
  select(-.estimator, -.config) |>
  pivot_wider(names_from = .metric, values_from = .estimate) |>
  # Excluded samples
  mutate(exclude = (n_occ < min_n_occ) & 
           !taxon %in% c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")) |>
  # Prepare table
  select(taxon, taxon_aliases, n_archaeo, n_occ, exclude, accuracy, roc_auc, 
         sens, p_predicted) |>
  arrange(taxon)

results_summary |>
  select(-taxon_aliases) |>
  gt() |>
  cols_label_with(fn = str_to_sentence) |>
  cols_label(
    n_archaeo = "Arch.",
    n_occ = "Cur.",
    accuracy = "Acc.",
    roc_auc = "ROC-AUC",
    sens = "Sens. cur.",
    p_predicted = "Sens. arch."
  ) |>
  tab_spanner("Occurrences", columns = c(n_archaeo, n_occ)) |>
  tab_spanner("Model", columns = accuracy:p_predicted) |>
  cols_hide(exclude) |>
  tab_footnote(
    "Excluded from modelling due to sample size", 
    cells_body(columns = c(n_occ), rows = exclude),
    placement = "right"
  ) |>
  add_taxon_alias_footnotes(results_summary$taxon_aliases) |>
  cols_align("right") |>
  cols_align("left", c(taxon)) |>
  fmt_number(accuracy:p_predicted) |>
  sub_missing() |>
  tab_style(cell_text(style = "italic"), cells_body(taxon)) |>
  tab_options(
    # footnotes.marks = "*",
    table.font.size = 11,
    table.width = pct(100),
    latex.use_longtable = TRUE
  )
```

@@

```{r fig-palaeodist-sum}
#| fig-cap: Predicted species richness (sum of predicted ranges) by period
# TODO: expand period facet labels
ggplot() +
  facet_wrap(vars(factor(period, levels = climate_periods$code, labels = climate_periods$label))) +
  annotation_spatial(ne_land, fill = "white") +
  geom_stars(data = select(palaeodist_sum, present), 
             na.action = na.omit) +
  annotation_spatial(ne_land, fill = NA) +
  annotation_spatial(w_asia, fill = NA) +
  scale_fill_batlow(reverse = TRUE) +
  coord_sf(crs = w_asia_albers) +
  labs(fill = "Predicted\nrichness", x = NULL, y = NULL) +
  theme(
    strip.background = element_blank(),
    strip.text.y = element_text(face = "italic")
  )
```

@@

Cyprus and the Aegean are not conventionally considered part of the primary zone of domestication but rather amongst the first regions that acquired agriculture from West Asia.
Our analysis complicates this picture, as it indicates that the wild ranges of many crop progenitors included these regions.
Early examples of several domesticates are recorded at sites on Cyprus, Western Anatolia and Greece [@ArranzOtaeguiRoe2023],
and the Aegean region was probably connected to West Asia by a land bridge via Anatolia until the Early Holocene [@AksuHiscott2022].
Were these area part of the same broader 'interaction sphere' that produced Neolithic agriculture in West Asia?

@@

Almost all the cereal and legume crop progenitors we modelled are predicted to have only been found in the Levant during the terminal Pleistocene and Early Holocene (see appendix).
Part of this may be do with the fact that both our initial flora and training occurrence dataset have a strong bias towards the southern Levant, but strikingly the modelled *current* ranges of these plants do tend to include Anatolia and the Zagros, so this cannot be the only factor.
One notable exception is the wild ancestor of chickpea (*Cicer reticulatum*), which is predicted to have a distribution centered on Northern Mesopotamia but encompassing much of the the 'hilly flanks' (except the southern Levant, @fig-palaeodist-chickpea-rye). 
Another is rye (*Secale cereale*), which is inferred to be primarily Anatolian (@fig-palaeodist-chickpea-rye).
This is perhaps relevant to rye's unusual domestication history, as a crop of West Asian origin that was intensively exploited [@HillmanEtAl2001; @DoucheWillcox2018] but apparently not first cultivated until much later than the 'founder crops', in Europe [@SchreiberEtAl2021].

@@

```{r fig-palaeodist-barely-levant, fig.height=10}
#| fig-cap: Predicted palaeodistribution of wild barleys in the Levant
# TODO: why OOM?`
#palaeodist_hordeum <- do.call(c, c(
#  filter(flora, str_starts(taxon, "Hordeum"))$palaeodist,
#  list(along = list(taxon = filter(flora, str_starts(taxon, "Hordeum"))$taxon))
#))
#
#ggplot() +
#  facet_grid(taxon ~ fct_relabel(period, toupper)) +
#  annotation_spatial(ne_land, fill = "white") +
#  geom_stars(data = select(palaeodist_hordeum, prediction.TRUE.), 
#             na.action = na.omit) +
#  annotation_spatial(ne_land, fill = NA) +
#  scale_x_continuous(breaks = c(34, 36, 38)) +
#  scale_fill_batlow(reverse = TRUE) +
#  coord_sf(
#    xlim = c(33, 39), ylim = c(30, 38),
#    crs = w_asia_albers, default_crs = latlong
#  ) +
#  labs(fill = "P(present)", x = NULL, y = NULL) +
#  theme(
#    strip.background = element_blank(),
#    strip.text.y = element_text(face = "italic")
#  )
```

@@

```{r fig-palaeodist-bread}
#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
bread <- c("Aegilops tauschii", "Triticum aestivum", "Triticum turgidum")
palaeodist_bread <- do.call(c, c(
  filter(flora, taxon %in% bread)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
)) |>
  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
  mutate(present = factor(present))

@@

## Disagreement between hindcast models and archaeobotanical composition {#sec-discuss-hindcasting}

In general there is middling to low agreement between our hindcast predictions of species niches and observed archaeological occurrences of the same species [@tlb-results-summary].

The majority of our models perform poorly as a binary classification with a threshold of `P(present)>0.5`.
However, this is an arbitrary threshold; lowering it can improve sensitivity at the cost of specificity.
<!-- TODO: so why didn't we threshold? Losing information - stick with the probabilities. -->
In principle archaeological occurrences—as true presences—could be used to 'calibrate' the threshold, but there aren't enough archaeological attestations of most of the modelled species.

<!-- TODO: What about not binary? Mean P(T) and P(F) for example? -->

```{r fig-archaeo-sensitivity}
#| fig-cap: Summary of hindcasting performance of models in relation to model performance and available archaeological test data
results_summary |>
  pivot_longer(c(n_archaeo, sens, n_occ), names_to = "metric") |>
  mutate(
    metric = case_match(
      metric,
      "n_archaeo" ~ "N occurrences (archaeological)",
      "n_occ" ~ "N occurrences (current)",
      "sens" ~ "Model sensitivity (current)"
    )
  ) |>
  ggplot(aes(value, p_predicted)) +
  facet_wrap(vars(metric), ncol = 1, scales = "free_x", 
              strip.position = "bottom") +
  geom_smooth(method = "lm", fullrange = TRUE, se = FALSE) +
  geom_point() +
  scale_x_continuous(labels = label_number()) +
  scale_y_continuous(limits = c(0, 1), labels = label_percent()) +
  labs(x = NULL, y = "Archaeological occurrences predicted") +
  theme_minimal_grid() +
  theme(strip.placement = "outside")
```

At this point we cannot explain precisely why this discrepency exists, though assessments of model performance offer some clues.

The hindcasting performance of models is not strongly correlated with either the number of occurrences in the training dataset, the performance of the model on test data, or the number of attested archaeological occurrences [@fig-archaeo-sensitivity].
This indicates against a single, simple explanation for the discrepancy, i.e. that one or the other signal is 'wrong'.

<!-- TODO: core issue - fundamental and realised niches, different but related signals -->

For a variety of reasons, our models probably underestimate the *fundamental* niche or potential distribution of the target taxon.
These include:

* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)
* Use of time-averaged climate slices

At the same time, there are several reasons to believe that the archaeobotanical data (at least as it has been compiled for this study) *overestimates* the fundamental niche:

* Time-averaged, sometimes quite severely
* (Understudied) possibility for human transplanting and transport; we know this is the dawn of agriculture!
* Generally small sample sizes, so individual misidentifications have a large impact

So what can we do about it???

<!-- TODO: rewrite following reviews -->
The failure of our hindcast models to predict the occurrence of species in archaeobotanical assemblages has several possible explanations. 
Since they do accurately predict the test dataset<!-- TODO: and past negatives? -->, a likely culprit is overfitting of the models to the present environment.
This implies that the modelled palaeodistributions should be seen as conservative estimates or a minimal range.
Another obvious flaw in our methodology is that the time slices used for palaeoclimatic reconstruction are very broad—each covering around two millennia—and therefore potentially unrepresentative of the environment around sites at the specific time at which they were occupied.
The variable quality of the archaeological test dataset, especially in terms of chronology, is also a plausible factor.

@@

"All models are wrong" [@Box1976], but the performance of our models on independent test datasets give confidence in its predictions, which are plausible and consistent with broad-scale patterns in the archaeological record.
Results with regard to specific archaeologically-attested species are not as promising, but this also reflects the incomplete and coarsely temporally-resolved nature of the archaeobotanical verification dataset.
Discrepancies between the modelled distribution, whether on the broad scale (e.g. the restricted geographic range of most species compared to their attestation in the archaeological record), or relating to specific species (i.e. false positives and false negatives), suggest several avenues for future investigation.
